---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::knit_hooks$set(
   error = function(x, options) {
     paste('\n\n

', sep = '\n')
   },
   warning = function(x, options) {
     paste('\n\n

', sep = '\n')
   },
   message = function(x, options) {
     paste('\n\n

', sep = '\n')
   }
)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# ecodiag

The `ecodiag` package allows to build ecological Diagnostic Tools (DT) in a standardized way. A DT corresponds to a set of random forest models (built using the `ranger` and `mlr` packages) designed to predict the probability of an ecological community being impacted by different driving factors (one model per driving factor). 

Still in development.

## Installation

You can install ecodiag from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("CedricMondy/ecodiag")
```

## Example

For examples, see the vignette `workflow`.
